---
title: "Examples for common file prep, plotting, anova, and regression"
author: "Michelle Voss"
output:
  github_document:
    toc: true
    toc_depth: 1
    html_preview: false

---


# **Clear previous, load packages**

```{r, warning=FALSE, include=FALSE}

rm(list=ls(all=TRUE)) 

# Load basic packages for working with data, we'll load others later too
library(knitr)
library(reshape2)
library(tidyverse)
library(stargazer)

R.Version()
getwd()
```



**This is a "wide format" file with data from our acute exercise manipulation in the bike project**  
* There are 34 participants, 11 variables

```{r}

# load acute variables data file
data = read.table("exercise_fileprep-plots-anova_data/acute_physio-variables-forR.csv", header=TRUE, sep = ",", fill = TRUE)

# look at features of your variables (encoded as integer, numeric or factors?)
str(data)

```



# **Referring to variables in a dataset**

```{r}

# change hr to percent for clarity of axes
# generic call of variable is dataframe$variable
# read <- as "is given by"
# here we are recoding, rather than making a new variable

data$hr_active <- 100*data$hr_active 
data$hr_passive <- 100*data$hr_passive

```


# **Reshaping from wide to long**
* Especially for longitudinal data, long format is much "tidyer"  
  * One nugget of info fits in each cell and use columns to label attributes of variable values  
  
  
```{r}

# reshape is a function to help transform from wide to long 
# look at the dataframe to see if you can tell how it's working
# it will not work if the variable name has two delimiters (e.g. notice all variables have only one underscore)
# when you name your variables initially, it's is very helpful to have the variable attribute (e.g., session, condition) following the underscore

data_long<-reshape(data,
                             varying=c(3:10),
                             timevar="condition",
                             direction="long",
                             idvar="bike_id",
                             sep="_") 

```



# **Plotting to understand the data**  
* label your chunk and any figures will be saved in a directory at low-res for rendering on github or sending to colleagues   
* Histograms and boxplots are nice for initial description  
  * Familiarize with distribution shape (normal or skewed) and anything funky (outliers)  



Simple box plot  
* aes <- aesthetics, basic parameters of what to plot from the dataframe  
* generic call: ggplot(dataframe,aes(x=variable-on-x-axis,y=variable-on-y-axis,fill=variable-to-color-code))  
* geom_boxplot() make a boxplot  
  * https://ggplot2.tidyverse.org/reference/geom_boxplot.html  
  * http://www.sthda.com/english/wiki/ggplot2-box-plot-quick-start-guide-r-software-and-data-visualization  
* geom_point() add points for individuals by making a scatter plot with a factor variable here  
  * https://ggplot2.tidyverse.org/reference/geom_point.html  

```{r}
ggplot(data_long, aes(x=condition,y=hr,fill=condition)) +
  geom_boxplot() +
  geom_point() +
  labs(title="Acute Phase Intensity",y="%HRmax",x="",fill="Condition") 
```



**GGplot is powerful and can help you add a lot of information to a plot like below**  

```{r acute-hr}

# center title
theme_update(plot.title = element_text(hjust = 0.5))

# ggplot tool
# each parentheses after the + adds a new feature to the plot

ggplot(data_long, aes(x=condition,y=hr,fill=condition)) + 
  scale_fill_manual(values=c("#E3871C", "#F5BD78")) +
  geom_boxplot() + 
  geom_point(aes(color=rpe)) +
  scale_color_gradient(low="gray",high="black",name="RPE") +
  labs(title="Acute Phase Intensity",y="%HRmax",x="",fill="Condition") +
  geom_text(aes(label=ifelse((hr>(quantile(hr,.75)+3*IQR(hr))|hr<(quantile(hr,.25)-3*IQR(hr))),paste(subject_id),"")), hjust=1.1, check_overlap = TRUE, size=3) + 
  theme(title=element_text(size=20, face='bold'),
        axis.text.x = element_text(size=20),
        axis.text.y = element_text(size=15),
        axis.title.y = element_text(size=20),
        legend.text = element_text(size=15),
        legend.title = element_text(size=15))

# save a high-resolution file in your directory for pub-level figures
ggsave("acute-hr.tiff", width = 200, height = 100, units = "mm",dpi=300) 

```



**How about a bar graph with error bars?**  
* fun.data=mean_se -> standard error  
* fun.data=mean_cl_normal -> confidence interval  
* Web resources  
  * http://murraylax.org/rtutorials/barplots.html  
  
  
```{r}

# center title
theme_update(plot.title = element_text(hjust = 0.5))

ggplot(data = data_long, aes(x = condition, y = hr)) +
  stat_summary(aes(y = hr), size=.5, fun.y = mean, geom="bar",size=1) + 
  stat_summary(fun.data=mean_cl_normal, geom="errorbar", size=.5,width=.3) + 
  labs(title="Acute Phase Intensity",y="%HRmax",x="",fill="Condition")  +
  scale_x_discrete(name="Condition") + 
  scale_y_continuous(name = "%HRmax") +
  ylim(0,100) +
  theme(title=element_text(size=20, face='bold'),
        axis.text.x = element_text(size=20,face='bold',angle = 30, hjust = 1),
        axis.text.y = element_text(size=20,face='bold'),
        axis.title.y = element_text(size=15))
```



**Show individuals on one plot**  

```{r}
ggplot(data_long, aes(x=condition,y=hr,group=bike_id,color=order)) + 
  geom_line() + 
  geom_point() +
  labs(title="Acute Phase Intensity",y="%HRmax")
```


**Power of facet wrap to plot by a factor**  
* by variable factor

```{r}

# add labels to a dummy coded factor
data_long$order<-factor(data_long$order,levels=c(1,0),labels=c("active-first","passive-first"))  


ggplot(data_long, aes(x=condition,y=hr,group=bike_id,color=order)) + 
  geom_line() + 
  geom_point() +
  facet_wrap(~order) +
  labs(title="Acute Phase Intensity",y="%HRmax") + 
  theme(strip.text = element_text(face="bold", size=15,lineheight=5.0), 
        strip.background = element_rect(colour="black", size=1))
```


* by subject  

```{r}
ggplot(data_long, aes(x=condition,y=hr,group=bike_id,color=order)) + 
  geom_line() + 
  geom_point() +
  facet_wrap(~bike_id)
```


# **Paired t-test**  
* nice little tutorial with graphing supplements
  * https://rcompanion.org/rcompanion/d_09.html
  
  
  
* for wide format

```{r}
t.test(data$hr_active, 
       data$hr_passive, 
       paired=TRUE, 
       conf.level=0.95)
```


* for long format
```{r}
t.test(hr ~ condition, 
       data=data_long, 
       paired=TRUE, 
       conf.level=0.95)
```


**ANOVAs**  
* nice little tutorial   
  * http://singmann.org/anova-in-r-afex-may-be-the-solution-you-are-looking-for/  

* long format

```{r warning=FALSE}

# does hr difference by condition also interact with order of sessions?
library(car)
library(afex)
```



```{r}
# repeated measures variables specified in error term by subject
# if you had no repeated meaures values you would write: Error(bike_id)
# if you had two repeated meaures values you would write: Error(bike_id/variable1*variable2)

aovHR<-aov_car(hr~condition*order+Error(bike_id/condition),data_long)
nice(aovHR)
aovHR_fitted<-lsmeans(aovHR,~condition|order)
aovHR_fitted
pairs(aovHR_fitted)

```




# **Regression with dummy coding**  
* nice little conceptual examples for coding predictors and implications for interpretation
  * http://faculty.cas.usf.edu/mbrannick/regression/anova1.html
  * http://www.lrdc.pitt.edu/maplelab/slides/Simple_Main_Effects_Fraundorf.pdf  

* use long format

Check data structure    
```{r}
str(data_long)
```

Change condition to a factor
```{r}
data_long$condition <-as.factor(data_long$condition)
str(data_long)
```


Check constrast coding
* notice "dummy coding" is default
```{r}
contrasts(data_long$condition)
contrasts(data_long$order)
```

With dummy coding, we typically want our control group or condition as the reference. This doesn't apply to order, but let's change the coding for condition so that passive is coded as 0.


```{r}
contrasts(data_long$condition)=c(1,0)
```

```{r}
contrasts(data_long$condition)
```

For reference, let's print our our cell means for the 2 x 2 of order*condition
* also see stargazer as a tool for outputting nice tables of descriptives that can be directly copied into a word doc  
  * https://www.jakeruss.com/cheatsheets/stargazer/

```{r}
library(psych)
describe(data_long)
describeBy(data_long,list(data_long$order,data_long$condition))
```

Resources comparing anova and regression   
* https://m-clark.github.io/docs/mixedModels/anovamixed.html
* mixed model generally: https://gkhajduk.github.io/2017-03-09-mixed-models/   


```{r}
# set bike_id as a factor since repeated measures are grouped by subject
data_long$bike_id <- as.factor(data_long$bike_id)

# the (1 | bike_id) text below helps account for the fact that there are repeated measures for subject
lm_hr_dummy<-lmer(hr ~ condition*order + (1|bike_id), data=data_long)
summary(lm_hr_dummy)
```

Below is an example of formatting the model results that is better for presentation. If you replace "text" with "html" then stargazer will save the table in an .html format, which you can then open in word to copy/paste the formatted table into your results tables.  
```{r}
class(lm_hr_dummy) <- "lmerMod"
stargazer(lm_hr_dummy,type="text",
          dep.var.labels=c("%HRR during exercise"),
          covariate.labels = c("Intercept","Condition","Order","Condition*Order"),
          star.cutoffs = c(0.05, 0.01, 0.001),
          intercept.bottom = FALSE)
```

Interpretations  
* What does the intercept represent?   
  * Generally, intercept is the mean of DV when all predictors are 0  
  * We dummy-coded our predictors...  
  * So intercept is mean HR when cond=0 (passive) and order=0 (active-first), match above? Yep (49.06)  
  
* Condition  
  * Has a 1 after label to indicate the parameter is for the contrast of [1 0], or here [active passive]  
    * This represents the mean difference of active-passive, ONLY AT THE REFERENCE for the order variable!   
      * reference: where order=0 (active-first)  

* Order  
  * label indicates passive-first coded as a 1  
  * As above, represents mean difference of order conditions, only where the other factor 0 (passive condition)  
    * Check this for yourself!  

* Interaction  
  * Generally, an interaction refers to when the difference in conditions for a variable depends on the level of another variable  
    * In words here, does the effect of condition on hr depend on which condition was first?   
  * Parameter reflects difference of differences: [level:passive-first](active-passive) -[level:active-first](active-passive)  
  * Interpretation: HR doesn't seem to depend on order, yay!  
  

* If we only care about the interaction, this is fine. But if we care about effect of condition regardless or order (e.g., a "main effect" of condition) we don't really that above like we did in the anova. We can change our coding scheme for factors to allow for this though.   


# **Regression with effect coding**   

For a two-level variable, effect coding assigns levels as 1, -1  


Remind ourselves of factor levels
```{r}
levels(data_long$condition)
levels(data_long$order)
```



Let's do this for both factors, where -.5 is our control condition if we have one
```{r}
contrasts(data_long$condition)=c(.5,-.5)
contrasts(data_long$order)=c(-.5,.5)
```

Check!
```{r}
contrasts(data_long$condition)
contrasts(data_long$order)
```

Run the same regression model, now with effect coded factors..
```{r}
lm_hr_effect<-lmer(hr ~ condition*order + (1|bike_id), data=data_long)
summary(lm_hr_effect)
```



Interpretations      
* What does the intercept represent?         
  * Generally, intercept is the mean of DV when all predictors are 0     
  * We effect-coded our predictors...        
  * So intercept is mean HR when cond=0 (passive) and order=0 (active-first), match above?     
  * That is now the grand mean across all conditions (60.13)    
    
* Condition       
  * Label indicates the parameter is for the contrast of [1 -1], or here [active passive]      
    * This represents the mean difference of active-passive, now at the MEAN of the order variable    
    * This is like our main effect of condition, regardless of order    

* Order    
  * label indicates contrast for [1 -1]    
  * As above, represents mean difference of order conditions, holding condition constant at its MEAN  

* Interaction    
  * Same as when we dummy coded...  
  



# **Continuous predictors** 

* Similar to rationale for effect coding, **if** the continous predictor is modeled to interact with the categorical predictors, it usually makes sense to mean-center a continous predictor so that other parameters reflect the effect of that variable at the mean-level of the continous predictor.  
  * If the continuous predictors are not interacting with the categorical predictors, they're treated more like covariates in parameters.  


Example, let's test if hr in the acute sessions is predicted by resting heart rate or cardiorespiratory fitness..   


bring these variables in from another .csv

```{r}
subvariables = read.table("exercise_fileprep-plots-anova_data/subvariables.csv", header=TRUE, sep = ",", fill = TRUE)
subvariables.k = subvariables %>% select(bike_id,rhr_m0,vo2rel_m0)

data_long <- merge(data_long,subvariables,all.x=TRUE, by="bike_id")

```



Resting heart rate as a covariate...  
```{r}
lm_hr_effect.c<-lmer(hr ~ condition*order+rhr_m0 + (1|bike_id), data=data_long)
summary(lm_hr_effect.c)
```

Resting heart rate as a covariate WITH mean-centering by z-score scaling...

```{r}
lm_hr_effect.c<-lmer(hr ~ condition*order+scale(rhr_m0) + (1|bike_id), data=data_long)
summary(lm_hr_effect.c)
```




Resting heart rate in the interaction...  
* translate what you saw with dummy vs effect coding, why doesn't this make any sense to do when looking at condition and order effects?     

```{r}
lm_hr_effect.c<-lmer(hr ~ condition*order*rhr_m0 + (1|bike_id), data=data_long)
summary(lm_hr_effect.c)
```

So this makes more sense if we want resting heart rate in the interaction

```{r}
lm_hr_effect.c<-lmer(hr ~ condition*order*scale(rhr_m0)+ (1|bike_id), data=data_long)
summary(lm_hr_effect.c)
```


Reporting regression results  

Often good to include a full table of beta parameters. Stargazer is a nice tool to help with this:
* https://www.jakeruss.com/cheatsheets/stargazer/ 
* https://www.princeton.edu/~otorres/NiceOutputR.pdf

```{r}

# use this for lmer models
class(lm_hr_effect.c) <- "lmerMod"

stargazer(lm_hr_effect.c,type="text",
          dep.var.labels=c("%HRR during exercise"),
          covariate.labels = c("Intercept", "Condition","Order","Resting HR (RHR)","Condition*Order","Condition*RHR","Order*RHR","Condition*Order*RHR"),
          star.cutoffs = c(0.05, 0.01, 0.001),
          intercept.bottom = FALSE)
```


Plot to understand interactions, nifty tool in package jtools  
* https://cran.r-project.org/web/packages/jtools/vignettes/interactions.html

```{r}
library(jtools)
```


```{r}
interact_plot(lm_hr_effect.c, pred = "rhr_m0", modx = "condition",plot.points = TRUE, data=data_long)
```


Curious about CRF ?    
* What would you predict?    
* Test it!   









